VariantSpark: Population Scale Clustering of Genotype Information

### 2015-09-22 doc/publications/journal2014/paper.tex
polished methods

\begin{keyword}
\kwd{sample}
\kwd{article}
\kwd{author}
\end{keyword}

@@

Genomic information is increasingly used in medical practice.
A commonly performed task in such applications is grouping individuals based on their genomic profile to identify population association~\cite{Gao2007} or elucidate different haplotype involvement in diseases susceptibility~\cite{Laitman2013}.  
The commonly used tool is {\sc Admixture}~\cite{Alexander2009}, which performs maximum likelihood estimation of individual ancestries from multilocus SNP genotype datasets.
Due to the decreasing sequencing cost it is economical to generate studies with sample sizes previously reserved for larger consortia such as the 1000 Genomes Project~\cite{1KG2012} or The Cancer Genome Atlas, TCGA~\cite{TCGA2013}. 
At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude. This drastic increase in both sample numbers and features per sample requires a massively parallel approach to data processing~\cite{Stein2010}. Traditional parallelisation strategies like OpenMPI or hardware accelerators (GPGPU) cannot scale with variable data sizes or require purpose-built hardware.

Addressing this issue, {\sc Apache Hadoop MapReduce}~\cite{Borthakur2007} transforms data into `key-value pairs' that can then be distributed between multiple nodes across a commodity computer cluster, depending on the size of the problem. 
MapReduce approaches are increasingly being used in bioinformatics (for reviews see~\cite{Zou2013, Qiu2010,Taylor2010}). 
This is especially the case for sequence analysis tasks, such as read mapping~\cite{Schatz2009}, duplicate removal~\cite{Jourdren2012}, and variant calling~\cite{Langmead2009, McKenna2010} as well as Genome Wide Analysis Study based tasks~\cite{Huang2013, Guo2014}. 
Apache also developed a machine learning library, Mahout~\cite{Owen2011}, which allows efficient out-of-the-box analysis to be applied to clinical applications, such as medical health records~\cite{Ko2014}.
Unfortunately, the MapReduce paradigm is not always the optimal solution, specifically for bioinformatics or machine learning applications that require iterative in-memory computation. Specifically, Hadoop is disk IO intensive, and this can prove to be a bottleneck in processing-speed.

{\sc Apache Spark}~\cite{Zaharia2011} is a more recent compute engine, which overcomes many of Hadoop's limitations. 
One of the main benefits is that it allows programs to cache data in memory; potentially eliminating, or at least reducing, the bottleneck of disk IO. 
When utilising caching, Apache claim {\sc Spark} to be 100x faster than Hadoop. 
Although {\sc Spark} allows MapReduce-like programs, it does not require programs to exactly model the MapReduce paradigm, which in turn allows more flexibility when designing programs. 
Recognising the capability, Wiewi{\'o}rka {\it et al.}~\cite{Wiewiorka2014} developed SparkSeq for high-throughput sequence data analysis and the Big Data Genomics (BDG) group recently demonstrated the strength of {\sc Spark} in a genomic clustering application using {\sc ADAM}, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}. 
While the speedup of {\sc ADAM} over traditional methods was impressive (50 fold speedup), being limited by constraints within this general genomics framework hampered performance.

We hence developed a more focused purpose-built application in {\sc Spark} to perform genomic clustering of individuals. 
We utilise {\sc Spark}'s machine learning library, \mbox{MLlib}, and provide an interface to the standard variant data format, Variant Call Format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms to a wide range of genotype-based analysis tasks.

To demonstrate \variantSpark's capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib. 
In the first section we benchmark \variantSpark's performance and accuracy against {\sc ADAM} and a Hadoop Mahout implementation as well as more traditional methods (R, Python) and the purpose-build tool {\sc Admixture}~\cite{Alexander2009} limiting the used data to chromosome 22.
In the second section we discuss the pipeline for visualising the resulting cluster.
In section three we demonstrate \variantSpark's full capacity by seamlessly scaling from 20\% to 100\% of the human genome.
In the last section we replicate the analysis by using 478 genomes from the Personal Genome Project~\cite{Lunshof2013}.

@@

As shown in Table 1, pre-processing the data is fastest in \variantSpark{}, requiring 2mins and 58secs. %178s 
This is almost 80\% faster than the {\sc ADAM} implementation (12min and 48secs) %768s
or our Hadoop implementation (14min and 22secs), respectively. %862s
Unlike \variantSpark{} and our Hadoop implementation, the {\sc ADAM} framework cannot process the VCF files directly but requires them to be converted into a binary {\sc ADAM} file format. 
Although this additional pre-processing step is only required once for each input file, it requires an additional 13mins and 13secs, %793s
rendering our approach almost an order of magnitude faster (see Figure 1).
R and Python are also slower at pre-processing, taking 67 minutes and 92 minutes, respectively. 
This increased pre-processing time is due to the Python and R implementations not natively supporting multithreading, while \variantSpark{}, {\sc ADAM} and Hadoop can use the eight available cores.
The GATK-based VCF to PED conversion is also slower with over 10 minutes.

Clustering the samples is also fastest in \variantSpark{} (1min and 20secs, see Table 1), despite \variantSpark{} and {\sc ADAM} both using {\sc Spark}'s MLlib \kMeans{} implementation. % 80
% DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
% AIDAN: This result was consistent. But it was on the Azure cluster so I can't get std/ste..
% DENIS: OK but I thought all of them were run on the same machine and you are still generating results for python ... 
% TODO:  write something like (repeated X times)
We attribute the 30\% speedup over {\sc ADAM} (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas {\sc ADAM} creates dense vectors, which are less memory efficient. %
Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise eight cores. 
Despite this, both are an order of magnitude slower than \variantSpark{} and {\sc ADAM}, with Python requiring 11mins and 29secs and R requiring 7mins and 25secs.
{\sc Admixture} is also slower with a runtime of just over 8 minutes. 
Our Hadoop implementation performs the worst with 14mins and 23secs, showcasing the limitations of the Hadoop engine. % 863
The substantially increased runtime is due to the dataset (chromosome 22) being small enough to fit into memory, which enables Python and R to store the output from each \kMeans{} iteration in memory. 
Hadoop, however, writes the entire output of each iteration to disk.
Although this Hadoop feature allows for greater scalability, as disk IO is not the run-time determining factor for larger datasets, it introduces a bottleneck for smaller datasets.
{\sc Spark} overcomes this limitation by performing in-memory caching of datasets, which eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can be cached in memory.

We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters. 
For this comparison, we use the \ARI{} (ARI) metric, which returns a value between -1 (independent labelings) and 1 (perfect match)~\cite{Hubert1985}.
Using chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
Even the state-of-the-art tool, {\sc Admixture}, does return a low ARI=0.25. It should be noted that {\sc Admixture}'s underlying statistical model does not take linkage disequilibrium (LD) into account. 
Therefore removing variants with high LD may result in higher accuracy.   
We can substantially improve the accuracy to a perfect classification (ARI=1.0) by removing the fourth super-population, AMR (American). 
This is due to the majority of AMR individuals being placed in the same group as Europeans, likely accurately reflecting their migrational backgrounds. 
Only a minority of AMR individuals form an independent group, likely comprising of genetic information otherwise not captured by the 26 sub-populations of the 1000 genomes project.
This indicates that although allelic differences exist between populations, "genetic diversity is distributed on a continuum"~\cite{Pugach2015}.

\subsection*{Graph visualisation}
\variantSpark{} also supports visualisation of the resulting population clusters by generating a file that can be loaded in Gephi \cite{ICWSM09154}.
Figure 3 visualises the clusters generated by \variantSpark{} from the 1000 Genomes Project data. 
As discussed in the previous section, three of the four clusters (super-populations AFR, EAS and AMR) are relatively-homogeneous; the fourth is more mixed and 
consists predominantly of individuals labelled EUR and AMR providing a visual representation of the ARI of 0.84.

\subsection*{\variantSpark{} allows genome wide sampling of variants to improve clustering quality}
Although we have demonstrated the advantages of \variantSpark{} over traditional methods on small datasets, the main advantage of {\sc Spark} is its ability to process datasets that exceed memory limits.
To demonstrate this scalability, we increase the number of variants used in the clustering from the initial 494,328 on chromosome 22 to 38,219,238 including all variants in the genome.

We run this comparison on our in-house Hadoop cluster using 40 cores.
Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively (see Table 2).
In each case, the memory required does not exceed a modest 2GB per executor, even for 100\% of the genome. 
Clustering the data takes 70 minutes, 139 minutes, 213 minutes and 884 mins, respectively.
For clustering, the memory requirements increase approximately linearly to 24GB per executor (see Figure 1). 
%TODO: doubleckeck
The increase in memory is due to more distance measurements between variants and \kMeans{} centroids being created, which MLlib stores as dense vectors. 
While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.

We do not observe an increase in cluster accuracy when providing variants from other chromosomes, indicating that current practice of clustering based on a subset of the genome is sufficient to define the existing population boundaries~\cite{Pugach2015}.

\subsection*{Personal Genome Project}
To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project (PGP). The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features~\cite{Lunshof2013}. 
We obtain and curate the data (see methods) resulting in 985,790 variants from 478 individuals.

@@

\variantSpark{} performs clustering on VCF files with over 1000 individuals and 40 Million variants in under 15 hours using minimal memory (24GB). 
\variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 85\% faster than Hadoop (29 min).
This novel parallelisation approach is also faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
\variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
These benefits of speed, resource consumption and scalability allows \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data.

\section*{Materials and methods}
\subsection*{Computational resources}
We completed the Chromosome 22 comparisons on a virtual machine (VM) hosted on Microsoft Azure. This VM is an A7 Linux instance with 8 cores, 56GB memory and Ubuntu as the OS. 
For the whole-genome clustering, we used our in-house Hadoop cluster. This uses Hadoop 2.5.0, managed by CDH 5. We use Spark 1.3.1. This 13 node cluster has a total of 416 cores and 1.22TB memory.

\subsection*{Datasets}
We used  phase 1 variants from the 1000 Genomes Project. On the 22 autosomes, this dataset contains variants for 1092 individuals, spread over a total 38,219,238 alleles.
We also used a subset of this dataset, the variants from chromosome 22, and this contains 494,328 alleles.
The individuals from these datasets are distributed across four super populations, African (AFR), Mixed American (AMR), East Asian (EAS) and European (EUR).

\subsection*{VariantSpark Implementation}
To cluster the individuals in a VCF file, we initially need to pre-process the variants to feature vectors.
In \variantSpark{} we read in the VCF files as text files to a Resilient Data Set (RDD). RDDs allow us to process the files, line-by-line, in parallel. We parse each line as tab-separated values, and `zip' each value with it's heading from the VCF file, as shown in Figure 2.
The result is an array of key-value pairs (KVPs) for each line, where the column heading is the key and the variant is the value.
We then `zip' each KVP array with a unique sequential index, where these indexes will serve as identifiers for the alleles.
Now each KVP array is the value in a parent KVP, where the key is the unique index.
We use a `flatMap' (step 3 of Figure 2) to convert these indexed arrays of KVPs into a less convoluted data-structure, as well as to filter out any extra metadata.
This results in KVPs (number of variants x number of alleles) where the key is the individual ID, and the value is a tuple of an allele and its index. Effectively, we have a KVP for each allele from the VCF file.
At this point, each allele is still a string straight from the VCF file (i.e. `0\textbar{}1'), however, we require doubles, for MLlib.
To convert these strings to doubles, we use the Hamming distance of the allele, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant. Also, for each individual, we filter out any alleles that have no variant, leaving us with KVPs of variants (homozygous and heterozygous) only.
We perform this `map' operation inside the previous `flatMap' (step 3).
We can now group these KVPs by their key (the individual ID) and save them to sparse vectors (step 4). This results in a sparse vector for each individual, the format required by MLlib's machine learning algorithms.

@@

\subsection*{Hadoop Implementation}
We use our previous Hadoop implementation, which is based on the MapReduce model. This makes use of KVPs, similarly to \variantSpark{}, however, it is explicitly based on the MapReduce model.
Because of this, and because we need a unique range of identifiers for alleles (in the smallest range possible), we need to run an initial MapReduce task to index the lines. This is comparable (however, more verbose) to the `zip' operation in \variantSpark{}.
The second MapReduce task does the bulk of the work and is visualised in Figure 2. The Map stage (step 1) begins by creating KVPs from the VCF file. For each KVP, the key is a tuple of a primary and secondary key, where the primary key is the individual ID and the secondary key is the allele ID. The value for each KVP is the variant.
The primary key ensures that KVPs for each individual are distributed to the same node during the MapReduce shuffle stage. After being distributed, the KVPs for each individual are sorted by their secondary key.
Now that the KVPs for each individual are physically located on the same hardware, the Reduce stage (step 2) can efficiently create a sparse vector for each individual from these KVPs. The sparse vectors (one for each individual) are saved to disk, ready to be used as the input to one of MLlib's algorithms.

%% Need to cite R package??
\subsection*{R Implementation}
For our R implementation, we read in a VCF file as a table and convert it to a matrix, dropping columns that don't represent alleles. As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value.
We then transpose the matrix, resulting in a data-structure where each row represents an individual. We convert the matrix to a `big.matrix', for the \kMeans{} algorithm from the `biganalytics' package.

\subsection*{Python Implementation}
Our Python implementation reads in lines from a VCF file as tab-separated values and stores the lines in a pandas DataFrame. The column headings are the individual IDs and the row headings are the allele locations.
We drop the first columns that aren't alleles, and then convert the remaining allele strings to numeric values. We can then transpose the DataFrame. Although to cluster this DataFrame (using sci-kit learn), we simply need to convert it to a matrix with {\sc .as\textunderscore{}matrix()}

\subsection*{{\sc Admixture} Implementation}
We use Genome Analysis Toolkit (GATK) to convert the variants from VCF, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
These three files are used as input to {\sc Admixture}, with $K$ (the number of ancestral populations) set to 4.

%% TODO - NEIL - (and better heading) %%
\subsection*{Personal Genome Project Data}
%% The PGP data..... 
PGP genome data was downloaded from the PGP website (\url{https://my.pgp-hms.org/public_genetic_data}). Genotype data from 23andMe microarray platforms was sorted by genome 
build. Shell scripts were written to convert NCBI build 36 files to BED format, update to build 37 using UCSC liftOver (\url{https://genome.ucsc.edu/cgi-bin/hgLiftOver}) and 
then convert back to 23andMe format. The 23andMe files were then converted to VCF format using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}). 
Where individuals had genotype data for both genome builds 36 and 37, only the latter was used. Finally, individual VCF files were combined for clustering into one 
file using VCFtools \cite{Danecek2011Variant}.

% DENIS yes, it will be in the git repository mentioned above (https://github.com/BauerLab/VariantSpark)
% can you please push your code into https://stash.csiro.au/users/bau04c/repos/variantspark/browse

@@

\section*{Figures}
  \begin{figure}[h!]
  \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.45]{images/Resources}
  \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.45]{images/Scaling}
  \caption{\csentence{Comparison of method and genome-wide scaling experiment.}
      {\bf Left} Runtime is given in seconds with 32 GB of memory on 8 threads (except for the pre-processing in R and Python where this was not supported). 
      {\bf Right} Scaling from 20\% to 100\% of variants in the genome with maximal number of executors and lowest possible memory assignment.}
      \end{figure}

>>>>

\begin{keyword}
\kwd{Genotype clustering}
\kwd{SPARK}
\kwd{BigData}
\kwd{1000 Genomes Project}
\kwd{Personal Genome Project}
\kwd{Population structure}
\end{keyword}

@@

Genomic information is increasingly used in medical practice.
A commonly performed task in such applications is grouping individuals based on their genomic profile to identify population association~\cite{Gao2007} or elucidate haplotype involvement in diseases susceptibility~\cite{Laitman2013}.  
The commonly used tool is {\sc Admixture}~\cite{Alexander2009}, which performs maximum likelihood estimation of individual ancestries from multilocus SNP genotype datasets.
Due to the decreasing sequencing cost it is now economical to generate studies with sample sizes previously reserved for larger consortia such as the 1000 Genomes Project~\cite{1KG2012} or The Cancer Genome Atlas, TCGA~\cite{TCGA2013}. 
At the same time, whole genome sequencing enables the inclusion of rare or even somatic mutations in the analysis, increasing the feature space by orders of magnitude. This drastic increase in both sample numbers and features per sample requires a massively parallel approach to data processing~\cite{Stein2010}. Traditional parallelisation strategies like MPI/OpenMP or hardware accelerators (GPGPU) cannot scale with variable data sizes at runtime~\cite{ReyesOrtiz2015121} or require purpose-built hardware.

Addressing this issue, {\sc Apache Hadoop MapReduce}~\cite{Borthakur2007} transforms data into `key-value pairs' that can then be distributed between multiple nodes across a commodity computer cluster, depending on the size of the problem. 
MapReduce approaches are increasingly being used in bioinformatics (for reviews see~\cite{Zou2013, Qiu2010,Taylor2010}). 
This is especially the case for sequence analysis tasks, such as read mapping~\cite{Schatz2009}, duplicate removal~\cite{Jourdren2012}, and variant calling~\cite{Langmead2009, McKenna2010} as well as Genome Wide Analysis Study based tasks~\cite{Huang2013, Guo2014}. 
Apache also developed a machine learning library, Mahout~\cite{Owen2011}, which allows efficient out-of-the-box analysis to be applied to clinical applications, such as medical health records~\cite{Ko2014}.
Unfortunately, the MapReduce paradigm is not always the optimal solution, specifically for bioinformatics or machine learning applications that require iterative in-memory computation. Specifically, Hadoop is relying extensively on hard disk input-output operations (disk IO), and this can prove to be a bottleneck in processing-speed.

{\sc Apache Spark}~\cite{Zaharia2011} is a more recent compute engine, which overcomes many of Hadoop's limitations. 
One of the main benefits is that it allows programs to cache data in memory; potentially eliminating, or at least reducing, the bottleneck of disk IO. 
When utilising caching, Apache claim {\sc Spark} to be up to 100x faster than Hadoop. 
Although {\sc Spark} allows MapReduce-like programs, it does not require programs to exactly model the MapReduce paradigm, which in turn allows more flexible software design. 
Recognising the capability, Wiewi{\'o}rka {\it et al.}~\cite{Wiewiorka2014} developed SparkSeq for high-throughput sequence data analysis and the Big Data Genomics (BDG) group recently demonstrated the strength of {\sc Spark} in a genomic clustering application using {\sc ADAM}, a set of formats and APIs as well as processing stage implementations for genomic data~\cite{Massie2013}. 
While the speedup of {\sc ADAM} over traditional methods was impressive (50 fold speedup), being limited by constraints within this general genomics framework hampered performance.

We hence developed a purpose-built application in {\sc Spark} to perform genomic clustering of individuals. 
We utilise {\sc Spark}'s machine learning library, \mbox{MLlib}, and provide an interface to the standard variant data format, Variant Call Format (VCF)~\cite{1KG2012}, which opens up the application of MLlib's different machine learning algorithms to a wide range of genotype-based analysis tasks.

To demonstrate \variantSpark's capability, we cluster variant datasets from the 1000 Genomes Project~\cite{1KG2012} to determine population structure using the \kMeans{} clustering algorithm available in MLlib. 
In the first section we benchmark \variantSpark's performance and accuracy against {\sc ADAM} and a Hadoop Mahout implementation as well as more traditional methods (R, Python) and the purpose-build tool {\sc Admixture}~\cite{Alexander2009}. % limiting the used data to chromosome 22.
In the second section we discuss the pipeline for visualising the resulting cluster.
In section three we demonstrate \variantSpark's full capacity by seamlessly scaling from 20\% to 100\% of the human genome.
In the last section we replicate the analysis by using 478 genomes from the Personal Genome Project~\cite{Lunshof2013}.

@@

As shown in Table 1, pre-processing the data is fastest in \variantSpark{}, requiring 2mins and 58secs. %178s 
This is almost 80\% faster than both the {\sc ADAM} implementation (12min and 48secs) %768s
and our Hadoop implementation (14min and 22secs). %862s
Unlike \variantSpark{} and our Hadoop implementation, the {\sc ADAM} framework cannot process the VCF files directly but requires them to be converted into a binary {\sc ADAM} file format. 
Although this additional pre-processing step is only required once for each input file, it requires an additional 13mins and 13secs, %793s
rendering our approach almost an order of magnitude faster (see Figure 1).
R and Python are also slower at pre-processing, taking 67 minutes and 92 minutes, respectively. 
This increased pre-processing time is due to the standard Python and R implementations not natively supporting multithreading, while \variantSpark{}, {\sc ADAM} and Hadoop can use the eight available cores.
The GATK-based VCF to PED conversion is also slower with over 10 minutes.

Clustering the samples is also fastest in \variantSpark{} (1min and 20secs, see Table 1), despite \variantSpark{} and {\sc ADAM} both using {\sc Spark}'s MLlib \kMeans{} implementation. % 80
We attribute the 30\% speedup over {\sc ADAM} (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas {\sc ADAM} creates dense vectors, which are less memory efficient. %
Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise eight cores. 
Despite this, both are an order of magnitude slower than \variantSpark{} and {\sc ADAM}, with Python requiring 11mins and 29secs and R requiring 7mins and 25secs.
{\sc Admixture} is also slower with a runtime of just over 8 minutes. 
Our Hadoop implementation performs the worst with 14mins and 23secs, showcasing the limitations of the Hadoop engine for iterative algorithms such as \kMeans{}. % 863
Hadoop writes the entire output of each \kMeans{} iteration to disk. This feature was designed to ensure scalability but here substantially increases the runtime as disk IO becomes the limiting factor for this small dataset (chromosome 22).
Python and R store the output from each iteration in memory thereby eliminating the IO bottleneck but without being able to utilise the disk they are limited to datasets that fit into memory.  
{\sc Spark} overcomes both limitations by performing in-memory caching for each executor, which eliminates the disk IO bottleneck on smaller datasets, while maintaining scalability to larger datasets where the data actively processed on executors is cashed in memory.

We also investigate the cluster quality for the five different methods by comparing the annotated super-population label (AMR, AFR, EAS, SAS) for each individual in the 1000 Genomes data to the assigned label from the clustering allowing four clusters. 
For this comparison, we use the \ARI{} (ARI) metric, which returns a value between -1 (independent labelings) and 1 (perfect match)~\cite{Hubert1985}.
Using chromosome 22, each of the algorithms resulted in an ARI of 0.84, confirming that the speed-up in {\sc Spark} is not at the cost of quality.
The state-of-the-art tool, {\sc Admixture}, returns a low ARI of 0.25. It should be noted that {\sc Admixture}'s underlying statistical model does not take linkage disequilibrium (LD) into account. 
Therefore removing variants with high LD may result in higher accuracy.   
We can substantially improve the accuracy to a perfect classification (ARI=1.0) by removing the fourth super-population, AMR (American). 
This is due to the majority of AMR individuals being placed in the same group as Europeans, likely reflecting their migrational backgrounds. 
Only a minority of AMR individuals form an independent group, likely comprising of genetic information otherwise not captured by the 26 sub-populations of the 1000 genomes project.
This indicates that although allelic differences exist between populations, "genetic diversity is distributed on a continuum"~\cite{Pugach2015}.

\subsection*{Graph visualisation}
\variantSpark{} supports visualisation of the resulting population clusters by generating a file that can be loaded into {\sc Gephi}~\cite{ICWSM09154}.
Figure 3 visualises the clusters generated by \variantSpark{} from the 1000 Genomes Project data. 
As discussed in the previous section, three of the four clusters (super-populations AFR, EAS and AMR) are relatively-homogeneous; the fourth is more mixed and 
consists predominantly of individuals labelled EUR and AMR providing a visual representation of the ARI of 0.84.

\subsection*{\variantSpark{} enables genome wide sampling of variants to improve clustering quality}
Although we have demonstrated the advantages of \variantSpark{} over traditional methods on small datasets, such as individual chromosomes, the main advantage of {\sc Spark} is its ability to process datasets that exceed memory limits.
To demonstrate this scalability, we increase the number of variants used in the clustering from the initial 494,328 on chromosome 22 to 38,219,238 including all variants in the genome.

We run this comparison on our in-house Hadoop cluster using 40 cores.
Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively (see Table 2).
In each case, the memory required does not exceed a modest 2GB per executor, even for the complete genome. 
Clustering the data takes 70 minutes, 139 minutes, 213 minutes and 884 mins, respectively.
For clustering, the memory requirements increase approximately linearly to 24GB per executor (see Figure 1). 
The increase in memory is due to more distance measurements between variants and \kMeans{} centroids being created, which MLlib stores as dense vectors. 
While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.

We do not observe an increase in cluster accuracy when providing variants from all chromosomes, indicating that current practice of clustering based on a subset of the genome is sufficient to define the existing population boundaries for populations with large genetic distance~\cite{Pugach2015}.

\subsection*{Determining the population structure of the Personal Genome Project}
To demonstrate the versatility of \variantSpark{} we also process and cluster individuals from the Personal Genomics Project (PGP). The PGP is open to the public for individuals to submit their genomic sequence along with any metadata, such as diseases or clinical features~\cite{Lunshof2013}. 
We obtain and curate the data (see methods) resulting in 985,790 variants from 478 individuals.

@@

\variantSpark{} performs clustering on VCF files with over 1000 individuals and 40 million variants in under 15 hours using minimal memory (24GB). 
\variantSpark{} supports random genome-wide sampling of variants allowing faster clustering for well characterised cohorts where 20\% of the genome is sufficient. 
On the benchmarking dataset, it outperforms {\sc ADAM} by almost an order of magnitude (4 vs 28 minutes) by processing VCF files directly and storing the information in sparse vectors. % 258 vs 1673 
\variantSpark{} utilises {\sc Spark}, which allows for in-memory caching and hence performs 85\% faster than Hadoop (29 min).
This novel parallelisation approach is also faster than traditional multithreading with 94\% speedup over R (75 min), and 95\% over Python (103 min).
\variantSpark{} is also superior in performance and accuracy over the current state-of-the-art tool for individual ancestries determination, {\sc Admixture}.
These benefits of speed, resource consumption and scalability allows \variantSpark{} to be the interface for applying other machine learning tools from MlLib to genomic data.

\section*{Materials and methods}
\subsection*{Computational resources}
We completed the Chromosome 22 comparisons on a virtual machine (VM) hosted on Microsoft Azure. This VM is an A7 Linux instance with 8 cores, 56GB memory running Ubuntu. 
For the whole-genome clustering, we used our in-house Hadoop cluster with Hadoop 2.5.0, managed by cloudera's CDH 5. We use Spark 1.3.1. This 13 node cluster has a total of 416 cores and 1.22TB memory.

\subsection*{Datasets}
We used  phase 1 variants from the 1000 Genomes Project. On the 22 autosomes, this dataset contains variants for 1092 individuals, spread over a total 38,219,238 alleles.
We also used a subset of this dataset, the variants from chromosome 22 containing 494,328 alleles.
The individuals from these datasets are distributed across four super populations, African (AFR), Mixed American (AMR), East Asian (EAS) and European (EUR).

\subsection*{VariantSpark Implementation}
To cluster the individuals in a VCF file, we initially need to pre-process the variants to feature vectors.
In \variantSpark{} we read in the VCF files as text files to a Resilient Data Set (RDD). RDDs allow us to process the files, line-by-line, in parallel. We parse each line as tab-separated values, and Spark's `zip' each value with it's heading (Sample name) from the VCF file, resulting in key-value pairs (KVPs) for each line see Figure 2 (Spark step 1). 
We then `zip' each pair with a unique sequential index, which later-on will serve as identifier for each allele (and keep the genome loci order), resulting in a KVP array (step 2).
We use `flatMap' to transpose the data structure to facilitate the grouping of variants by individuals.
For the \kMeans{} clustering, we convert the VCF genotypes (i.e. `0\textbar{}1') to doubles using the Hamming distance to the reference, where a 1 represents a heterozygous variant, 2 a homozygous variant, and 0, no variant. 
To remove uninformative KVPs,  we filter out any alleles that have no variant. 
We perform this `map' operation inside the previous `flatMap' (step 3).
We can now group these KVPs by their key (the individual ID) and save them to sparse vectors (step 4).

@@

\subsection*{Hadoop Implementation}
Our Hadoop implementation is based on the MapReduce model, which utilises KVPs, similarly to \variantSpark{}.
As we need a unique range of identifiers for alleles (in the smallest range possible), we need to run an initial MapReduce task to index the lines. This is comparable (however, more verbose) to the `zip' operation in \variantSpark{}.
The second MapReduce task does the bulk of the work and is visualised in Figure 2. The Map stage (step 1) begins by creating KVPs from the VCF file. For each KVP, the key is a tuple of a primary and secondary key, where the primary key is the individual ID and the secondary key is the allele ID. The value for each KVP is the variant.
The primary key ensures that KVPs for each individual are distributed to the same node during the MapReduce shuffle stage. After being distributed, the KVPs for each individual are sorted by their secondary key.
Now that the KVPs for each individual are physically located on the same hardware, the Reduce stage (step 2) can efficiently create a sparse vector for each individual from these KVPs.

\subsection*{R Implementation}
For our R implementation, we read in a VCF file as a table and convert it to a matrix, filtering columns where all variant as conform to the reference. As with our \variantSpark{} pre-processing, we convert the strings that represent each allele to a numeric value.
We then transpose the matrix, resulting in a data-structure where each row represents an individual. We convert the matrix to a `big.matrix', for the \kMeans{} algorithm from the `biganalytics' package (\url{https://cran.r-project.org/web/packages/biganalytics/index.html}).

\subsection*{Python Implementation}
Our Python implementation reads in lines from a VCF file as tab-separated values and stores the lines in a pandas DataFrame (\url{http://pandas.pydata.org}). The column headings are the individual IDs and the row headings are the allele locations.
We filter out reference-conform data and convert the remaining allele strings to numeric values. 
We cluster this DataFrame using sci-kit learn (\url{http://scikit-learn.org/stable/}), upon converting it to a matrix with {\sc .as\textunderscore{}matrix()}.

\subsection*{{\sc Admixture} Implementation}
We use Genome Analysis Toolkit ({\sc -T VariantsToBinaryPed})~\cite{McKenna2010} to convert the variants from VCF, and a 1000 Genomes Project supplied .ped file, to a binary PLINK (.bed), binary marker information file (.bim) and pedigree stub file (.fam).
These three files are used as input to {\sc Admixture}, with $K$ (the number of ancestral populations) set to 4.

\subsection*{Personal Genome Project Data}
Genome data was downloaded from the Personal Genome Project Data (PGP) website (\url{https://my.pgp-hms.org/public_genetic_data}). Genotype data from 23andMe microarray platforms was sorted by genome 
build. Shell scripts were written to convert NCBI build 36 files to BED format, update to build 37 using UCSC liftOver (\url{https://genome.ucsc.edu/cgi-bin/hgLiftOver}) and 
then convert back to 23andMe format. The 23andMe files were then converted to VCF format using code obtained from the Broad Institute (\url{http://apol1.blogspot.com.au/2013/08/impute-apoe-and-apol1-with-23andme.html}). 
Where individuals had genotype data for both genome builds 36 and 37, only the latter was used. Finally, individual VCF files were combined for clustering into one 
file using VCFtools \cite{Danecek2011Variant}.

@@

\section*{Figures}
  \begin{figure}[h!]
  \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.45]{images/Resources}
  \includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.45]{images/Scaling}
  \caption{\csentence{Comparison of method and genome-wide scaling experiment.}
      {\bf Left} Runtime for clustering variants from chromosome 22 is given in seconds with 32 GB of memory on 8 threads (except for the pre-processing in R and Python where this was not supported). 
      {\bf Right} Scaling from 20\% to 100\% of variants in the genome with maximal number of executors and lowest possible memory assignment.}
      \end{figure}

\begin{figure}[h!]
\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.25]{signature}
  \caption{\csentence{Schematic of the Hadoop and Spark implementation.} The image shows the conceptual differences in the steps necessary to convert variants in VCF to a data structure readable by Mahout and MLlib, respectively.}
      \end{figure}

\begin{figure}[h!]
\includegraphics[type=pdf,ext=.pdf,read=.pdf, scale=0.40]{chr22-k4-gephi-legend}
  \caption{\csentence{Visualisation of \variantSpark{} predicted clusters.} Individuals from the 1000 Genomes project are clustered effectively by super-populations AFR, AMR and EAS; a fourth cluster is predominantly EUR + AMR.}
      \end{figure}
